---
title: "A view of spiroview"
author: "Annie Liu"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Introduction_spiroview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.align = "center", 
               out.width = "90%",
               fig.width = 6, fig.height = 5.5,
               dev.args=list(pointsize=10),
               par = TRUE, # needed for setting hook 
               collapse = TRUE, # collapse input & ouput code in chunks
               warning = FALSE)
knit_hooks$set(par = function(before, options, envir)
  { if(before && options$fig.show != "none") 
       par(family = "sans", mar=c(4.1,4.1,1.1,1.1), mgp=c(3,1,0), tcl=-0.5)
})
set.seed(1) # for exact reproducibility

library("spiroview")
```


## Introduction
`spiroview` is an R package developed to provide the tools to allow for easy exploration and visualization of spirometry data in demographic studies. This package is targeted for researchers interested in answering questions in the field of respirology using spirometry but may not have an extensive coding background. The functions in `spiroview` allow for data extraction, calculations of expected spirometric values, and simple data visualization. 

**This document gives a tour of spiroview (version 0.1.0)**. It was written in R Markdown, using the [knitr](https://cran.r-project.org/package=knitr) package for production. See `help(package = "spiroview")` for further details. 

``` r
require("devtools")
devtools::install_github("annie951977/spiroview", build_vignettes = TRUE)
library("spiroview")
```

To list all sample functions available in the package:
``` r
ls("package:spiroview")
```

To list all sample datasets available in the package:
``` r
data(package = "spiroview")
```

<br>


## Brief Tutorial 

Here, we give a brief tutorial on how to use `spiroview` with the provided dataset.

``` r
exampleDB <- GLIData
```

### Reading in a dataset 
Alternatively, a user may load a dataset.

If your dataset already follows the conventions defined in `formatData`, a user 
may choose to simply read in a dataset with the following function.

``` r
# path to the dataset
examplePath <- system.file("extdata",
                           "example_dataset_1.csv",
                            package="spiroview")
                            
                            
readInDB <- read.csv(exmaplePath, headers=TRUE)

head(readInDB)
```

Otherwise, a user may choose to use `formatData` to format key headers for other
functions. Here, we format an example dataset, passing in the column name for age
that already existed in the dataset.

``` r

examplePath <- system.file("extdata",
                           "example_dataset_4.csv",
                            package="spiroview")


formattedDB <- spiroview::formatData(path=examplePath,
                                    ageLabel= "A")
                                    
head(formattedDB)
```

See `?spiroview::formatData` for more information on how to use this function.


### Calculating predicted spirometry values and summary statistics

To calculate a column for the predicted lower limit of normal for each row in
your dataset, call `calculateLLNPret` and input the spirometric parameter and 
reference equation of interest.

``` r

 predictedFEV1LLN <- calculateLLNPret(
                     df=exampleDB,
                     param="FEV1",
                     ref="GLI")
```

To add the new predicted column, use the base R function `cbind` which takes a 
existing dataframe and adds another dataframe as a column.

```
 cbind(exampleDB, predictedFEV1LLN)
```

To summarize data of patients meeting a certain threshold:

```
 categoricalSummary <- summarizeAllByCategory(df=exampleDB,
                                             demParam="gender",
                                             delim="1",
                                             delimIsNumeric = FALSE,
                                             spiroParam="FEV1")
                                             
 categoricalSummary

```

### Segregating Dataset

To segregate by a categorical variable, add the name of the variable and the 
value of the variable you wish to extract. 

``` r
 segregateCategorical <- segregateBy(df=exampleDB,
                                      demParam = "gender",
                                      segBy = "1",
                                      segIsNumeric = FALSE)

 segregateCategorical$contains # dataframe that matches demographic threshold
 
 segregateCategorical$other # dataframe not matching thresold
```

To segregate by a numerical variable, add the name of the variable set a rule
for what value of that numerical variable you wish to extract. Set the parameter
`segIsNumeric` to `TRUE`.

``` r
 segregateNumerical <- segregateBy(df=exampleDB,
                                    demParam = "height",
                                    segBy = ">1.80",
                                    segIsNumeric = TRUE)

 segregateNumerical$contains # dataframe that matches demographic threshold
 
 segregateNumerical$other # dataframe not matching thresold
```

### Visualization of Data

To get a basic graph comparing a categorical variable to a spirometric variable:

``` r
 categoricalPlot <- viewCategorical(df=exampleDB,
                                   demParam="ethnicity",
                                   spiroParam="FEV1",
                                   type="box",
                                   title="Ethinicity to FEV1")
 

 categoricalPlot
```

To view the counts of each categorical variable:

``` r
 countsPlot <- viewCategoricalCounts(df=exampleDB,
                                     demParam="ethnicity",
                                     type="pie")
 countsPlot
```

To get a basic graph comparing a numerical variable to a spirometric variable:

```
 numericalPlot <- viewNumerical(df=exampleDB,
                               demParam="height",
                               spiroParam="FEV1",
                               includeBestFit=TRUE,
                               title="Height to FEV1")
                               
 numericalPlot
```

To produce a graph that annotates a numerical variable by a colour:
``` r
 compareNumericalPlot <- compareNumerical(df=exampleDB,
                                         demParam="height",
                                         delim=">1.60",
                                         spiroParam="FEV1",
                                         includeBestFit=TRUE,
                                         title="Height to FEV1")
                                        
 compareNumericalPlot
```
## Running Shiny 
To launch the Shiny app:

``` r
spiroview::runspiroview()
```

## Citing spiroview

Liu, A. (2023) spiroview: An R Package for the Exploration and Visualization of Demographic Spirometry Data.
  Unpublished. URL https://github.com/annie951977/spiroview
  
<br>

## References

- Chang W, Cheng J, Allaire J, Sievert C, Schloerke B, Xie Y, Allen J,
  McPherson J, Dipert A, Borges B (2023). shiny: Web Application
  Framework for R. R package version 1.7.4.9002,
  <https://shiny.rstudio.com/>.

- Cooper, B. G., Stocks, J., Hall, G. L., Culver, B., Steenbruggen, I.,
  Carter, K. W., Thompson, B. R., Graham, B. L., Miller, M. R., Ruppel,
  G., Henderson, J., Vaz Fragoso, C. A., & Stanojevic, S. (2017).The
  Global Lung Function Initiative (GLI) Network: bringing the world’s
  respiratory reference values together. Breathe (Sheffield, England),
  13(3), e56–e64. <https://doi.org/10.1183/20734735.012717>

- Grolemund, G. (2015). Learn Shiny - Video Tutorials.
  URL:<https://shiny.rstudio.com/tutorial/>

- Hankinson, J. L., Odencrantz, J. R., & Fedan, K. B. (1999).
  Spirometric reference values from a sample of the general U.S.
  population. American journal of respiratory and critical care
  medicine, 159(1), 179–187.https://doi.org/10.1164/ajrccm.159.1.9712108

- Lytras T (2020). *rspiro: Implementation of Spirometry Equations*. R
  package version 0.2, <https://CRAN.R-project.org/package=rspiro>.

- R Core Team (2021). R: A language and environment for statistical
  computing. R Foundation for Statistical Computing, Vienna, Austria.
  URL <https://www.R-project.org/>.

- Silva, A. (2022) TestingPackage: An Example R Package For BCB410H.
  Unpublished. URL <https://github.com/anjalisilva/TestingPackage>.”

- Wickham, H. and Bryan, J. (2019). *R Packages* (2nd edition). Newton,
  Massachusetts: O’Reilly Media. <https://r-pkgs.org/>

- Wickham H, François R, Henry L, Müller K, Vaughan D (2023). dplyr: A
  Grammar of Data Manipulation. <https://dplyr.tidyverse.org>,
  <https://github.com/tidyverse/dplyr>.

- Wickham H, Hester J, Bryan J (2023). readr: Read Rectangular Text
  Data. <https://readr.tidyverse.org>,
  <https://github.com/tidyverse/readr>.

- Wickham H (2016). ggplot2: Elegant Graphics for Data Analysis.
  Springer-Verlag New York. ISBN 978-3-319-24277-4,
  <https://ggplot2.tidyverse.org>.

- Wickham H (2022). *stringr: Simple, Consistent Wrappers for Common
  String Operations*. R package version 1.5.0,
  <https://CRAN.R-project.org/package=stringr>.


``` {r}
sessionInfo()
```
